\mytitle

### 2015-08-06 MSc_diss.tex
Updates in model description

Alkyl nitrates also have a biomass burning source which have been first reported by \cite{Simpson2002}. They found that in the immediate vicinity of savanna fires in Northern Australia the average alkyl nitrate mixing ratios were 47-–122 times higher than local background mixing ratios, exceeding 3 ppbv for methyl nitrate. Methyl nitrate was primarily emitted during the 'flaming' stage, while $C_2$-$C_4$ ethyl-butyl nitrates dominated during the 'smoldering' stage.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\arraystretch}{1}
\begin{table}[h!]
\caption{!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!}\label{tab:ANmean} % mean AN mixing ratios
\begin{tabularx}{\linewidth}{P{3cm}P{2cm}P{5cm}X}
\hline
Environment     & Mean $\Sigma RONO_2$ & $\Sigma RONO_2$ breakdown & Reference \\
\hline
remote marine (North Atlantic), airborne   & 10 pptv    & methyl, ethyl, 1-propyl, 2-propyl, 2-butyl, 2-methyl-3-butyl, sum of 2- and 3-pentyl  & \citep{Reeves2007} \\
\hline
rural ground (Nova Scotia, Canada)        & 14 pptv    & methyl, ethyl, n-propyl, 2-propyl, 2-butyl  & \citep{Roberts1998} \\
\hline
suburban ground (Cape Retire, South Africa)  & 17.48 pptv (3.62-122.23) & 2-propyl, 1-propyl, 2-butyl, 1-butyl, 3-pentyl, 2-pentyl, 2-methyl-1-butyl, i-pentyl & \citep{DeKock1994} \\
\hline
suburban ground (Black Forest, Germany)    & 119.8 ppt (30-630 ppt) & $C_1-C_8$ & \citep{Flocke1998} \\
\hline
polluted marine airborne (Gulf of Mexico) & approx 1 ppb (plume center)& $NO_y-(NO_x+HNO_3+PAN)$ & \citep{Neuman2012} \\
\hline
\end{tabularx}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

A study that enabled us to focus on examination of $O_3$ and $RONO_2$ formation processes has been conducted by \cite{Neuman2012}. After the Deepwater Horizon (DWH) offshore oil platform accident on 20 April 2010 large amounts of oil from the DWH spill reached the ocean surface. Subsequent oil evaporation caused a release of little methane and other light alkanes to the atmosphere forming a $VOC$ mixture atypical for the remote marine environment. At the same the $NO_x$ levels were enhanced in the vicinity of the oil spill due to spill response efforts (ship exhaust, surface oil burning, flaring of recovered gas). Airborne measurements of reactive nitrogen species ($NO_y$ = $NO_x$+$HNO_3$+$PAN$\footnote{Peroxyacetyl nitrate.}+$RONO_2$) and VOCs in the marine boundary layer during daytime downwind from the DWH area showed that oxidation of alkanes with between 5 and 10 carbons per molecule ($C_5$–$C_{10}$) dominated $O_3$ formation chemistry in the plume. This circumstance led to a rapid and nearly complete oxidation of $NO_x$ to $RONO_2$ with mostly 6--10 carbons per molecule ($C_6$–$C_{10}$) effectively terminating $O_3$ production. On 10 June, $O_3$ enhancements in the DWH plumes were less than 15 ppbv, even after $NO_x$ was completely oxidized.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h!]
\centering
\includegraphics[width=0.5\linewidth]{{\picspath Neuman2012_Fig.4}.png}
\caption{Time series of reactive nitrogen species measured on 10 June 2010 in the MBL in plumes (a) 10 km and (b) 50 km downwind from DWH. $NO_x$, $NO_y$, and $HNO_3$ were measured once per s and $PAN$ was measured once per 2 s. $C_1$–$C_5$ alkyl nitrates were measured in whole air samples at locations indicated by the circles. The sum of alkyl nitrates (black lines) is inferred from the difference between $NO_y$ and ($NO_x$+$HNO_3$+$PAN$) \citep{Neuman2012}.}\label{fig:Neuman2012}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

$O_3$ production efficiencies in the DWH plume were 6 plus minus 0.5 ppbv $O_3$ per ppbv of $NO_x$ oxidized after 1 h of transport. These values are comparable with those measured in U.S. urban areas \textbf{[Trainer et al., 1995; Neuman et al., 2009] ADD THESE TO BIBTEX}, but smaller than those measured in the petrochemical plumes with $OH$ loss rates dominated by alkenes (10--18 ppbv $O_3$ per ppbv of $NO_x$) [Ryerson et al., 2003] AND THIS ONE!!!.

Temporal changes in reactive nitrogen partitioning in the DWH plume (Figure \ref{fig:Neuman2012}) nicely illustrate that in fresh part of the plume (Figure \ref{fig:Neuman2012}a) $NO_x$ accounted for over 80\% of plume $NO_y$, in aged plumes (Figure \ref{fig:Neuman2012}b), $NO_x$ oxidation was nearly complete, but NOx, PAN, HNO3, and the sum of eight $C_5$–$C_{10}$ $RONO_2$ accounted for only 22\% of plume $NO_y$. It means that $C_6$–$C_{10}$ $RONO_2$ account for 78\% of the aged plume $NO_y$ \citep{Neuman2012}. This result differs dramatically from previous studies, where the estimated contribution of $RONO_2$ to $NO_y$ was much lower: approximately 4\% in urban and petrochemical plumes [Neuman et al., 2002; Ryerson et al., 2003], approx. 10\% in rural areas [Parrish et al., 1993], a few percent in urban air masses [O’Brien et al., 1997; Flocke et al.,1998; Schneider et al., 1998; Talbot et al., 2003; Ryerson et al., 2003], and up to 10–20\% of NOy in some studies [Day et al., 2003; Perring et al., 2010] AND ALL OF THESE ABOVE!!!!!!!.

Probably the most detailed investigation of the formation processes and photochemical evolution of alkyl nitrates (but not their impact on $O_3$) has been done by \cite{Sommariva2008}. Using a subset of the Master Chemical Mechanism v3.1 (MCM) they constructed a highly detailed chemical model and run it for 5 days under two scenarios (Table \ref{tab:Sommariva2008init}) to explore the sensitivity of the evolution of the chemical composition of an air mass emitted from an urban environment to the initial conditions. As it seen in Figure \ref{fig:Sommariva2008fig} according to the modelling results $O_3$ is produced and $NO_x$ is removed from the system with the consequent formation of $HNO_3$ via the reaction of $NO_2$ with $OH$, which broadly agrees with the observations [(Neuman et al., 2006).]. In scenario A the concentration of $NO_x$ after the first day is about 0.5–-1.0 ppb for $NO_2$ and approximately 100 ppt for $NO$, which is high enough to sustain the production of $O_3$ to up to 135 ppb and the formation of alkyl nitrates during the simulation. In scenario B (Figure \ref{fig:Sommariva2008fig} right) the increase in $O_3$ is slower (note that initial concentration is 0) and the maximum concentration after five days is about 100 ppb. The concentration of $NO_x$ remains relatively high during the second (10–-15 ppb) and third day (3-–5 ppb) and then decreases to $<$200 ppt of $NO$ and $<$1ppb of $NO_2$ on the fourth and fifth days, therefore allowing the formation of alkyl nitrates too. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
\caption{Initial values of the some model parameters (in ppb) for two scenarios used in \citep{Sommariva2008}}\label{tab:Sommariva2008init}
\centering
\begin{tabular}{ccc}
\hline
Parameter & Initial value Scenario A & Scenario B \\
\hline
$CO$      & 332.5 & 232.5 \\
$NO_x$    & 34.9  & 46.5  \\
$O_3$     & 90    & 0   \\
\hline
\end{tabular}
\end{table}
\begin{figure}
\includegraphics[width=\linewidth]{{D:/UEA/ENV-MB4X_Dissertation/ENV-MB4X_Dissertation_thesis/ENV-MB4X_Dissertation_thesis_latex/pics/Sommariva2008_Fig.1}.png}
\caption{Modelled concentrations of $O_3$, $NO_x$ and $HNO_3$ for scenario A (left) and B (right). Source: \citep{Sommariva2008}.}\label{fig:Sommariva2008fig}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Comparison of modelling results with the values observed inside a typical urban plume sampled using the NOAA WP-3D aircraft from the New York City area during the New England Air Quality Study (NEAQS) 2004 campaign enabled \citep{Sommariva2008} to conclude that the chemistry of alkyl nitrates is well represented in the MCM model. They found that short-chain $<C_4$ alkyl nitrates can be formed from several precursors, while long-chain ($C_5$) alkyl nitrates are produced for 90\% or more from the oxidation of a single parent alkane. They also found that at lower $O_3$ and $CO$ initial concentrations and higher $NOx$ (scenario A) the importance of the reactions between alkane and $OH$ is typically lower during the first two-three days, and the contributions of peroxides and carboxylic acids become significant later and are on average less important.

The first 3-D chemical transport  model simulations examining the impact of methyl and ethyl nitrates on tropospheric ozone chemistry in remote regions have been conducted by \cite{Neu2008}. In their study, they incorporated methyl and ethyl nitrates in the chemical mechanism of the UCI CTM [Carver et al., 1997] and by matching their observed atmospheric abundances derived a total oceanic emission flux of these two nitrates (0.35 Tg of N per year). Comparing two simulations with and without spatially uniform emissions of methyl and ethyl nitrates they found that the largest impact of oceanic alkyl nitrates is seen over the equatorial Western Pacific (Figure \ref{fig:Neu2008}). Atmospheric circulation (convergence of the trade winds) in this region isolates the marine boundary layer, resulting in sustained photochemical loss of $O_3$ (Figure \ref{fig:Neu2008M_O3contol}) and at the same time accumulation of alkyl nitrates (Figure \ref{fig:Neu2008M_MeONO2}), which drive up boundary layer $NO_x$ abundances by as much as 250\% (Figure \ref{fig:Neu2008M_NOxdiff}). Such combination of factors leads to increases in $O_3$ of up to 20\% (Figures \ref{fig:Neu2008M_O3contol} and \ref{fig:Neu2008M_O3diff}), with the largest increases occurring from May to October.

@@

To sum up, alkyl nitrates are produced as a minor branch of the reaction of peroxy radicals with NO \citep{Day2003}. Since these compounds are removed from the troposphere by photolysis and OH oxidation they can provide information on photochemical evolution of an air mass through relationships with their parent hydrocarbons \citep{Bertman1995}. Acting as reservoir species for reactive nitrogen (NOy) and having a relatively long atmospheric lifetimes alkyl nitrates play an important role in long range transport of NOx. Furthermore, alkyl nitrates link to ozone and carbonyl production making them useful tracers for both species \citep{Worton2010}. Despite their relative importance alkyl nitrates are usually not included in the climate-chemistry models, for example in UKCA (Reeves, personal communication), which potentially could lead to discrepancies between models and observations described before.

\section{Methodology} \label{sec:method}
\subsection{Model development and description}\label{sec:method_develop}
A zero-dimensional (box) model is developed to investigate the impact of alkyl nitrate chemistry on ozone production efficiency. Its code is written using a new to the author program \href{http://www.mcpa-software.com/}{FACSIMILE for Windows}. FACSIMILE is a special high-level programming language which is a powerful modelling tool designed to solve differential equations including those resulting from a set of chemical reactions using the Gear’s method.

The model is intended to represent the steady state of the troposphere under mean summer daytime meteorology and clear skies. To achieve that a number of assumptions have been made. First, the model assumes a uniform mixing of individual constituents of the troposphere, second, there are no atmospheric transport (no 'winds') and no emissions of species into the box meaning that the time evolution of each species is controlled by the chemical interactions alone and determined by their initial concentrations. The air temperature and relative humidity are assumed to be $25^{\circ}C$ and 70\% respectively and are fixed in all model runs. Other assumptions concerning the chemical mechanism and photolysis rates can be found in the following sections.

@@

\item the formation of alkylperoxy radicals ($RO_2$) from oxidation of the hydrocarbons;

@@

The full list of reactions can be found in Appendix \ref{sec:appendix1}. It is important to mention that due to purely theoretical purpose of this research and its focus on investigation of the impact of alkyl nitrate chemistry on ozone production some of the reactions that lead to formation of species that serve as a sink for $NO_x$ are not included into the chemical mechanism. However, these reactions play an important role in the chemistry of the troposphere, and below here is a description of their role and anticipated effect of their exclusion.

@@

\item The formation of the nitric acid ($HNO_3$) is a major loss process for $NO_x$ during daytime in the atmosphere. Being a 'sticky' molecule $HNO_3$ readily adsorbs to surfaces, particularly if there is water on the surface, and deposits either in precipitation (wet deposition) or in dry form (dry deposition). Because of the high deposition velocity, dry deposition of $HNO_3$ can be responsible for much of the removal of inorganic nitrogen from the troposphere \citep{Finlayson-Pitts2000}, which means that exclusion of $HNO_3$ (formation and removal)/(chemistry) from the chemical mechanism potentially leads to overproduction of ozone in the model, especially at high $NO_2$ concentrations because the competition between $NO_2$ photolysis and its reaction with $OH$ radical ($NO_2 + OH \xrightarrow{M} HNO_3$) is ignored.
\item Since this research is focused on the daytime chemistry some reactions that are believed to be significant in nighttime are not included into the chemical mechanism of the model. Among them are reactions that involve the nitrate radical ($NO_3$) and dinitrogen pentoxide ($N_2O_5$).

@@

\item $NO_3$ radical rapidly photolyzes during the daylight hours, but in nighttime its concentration can increase to measurable levels \citep{Atkinson2008}. This increase in concentration and a lesser availability of the OH radical (which sources are mostly photolytic in nature) make $NO_3$ radical a major contributor to the night chemistry of organics in the troposphere \citep{Finlayson-Pitts2000}. However, exclusion of $NO_3$ radical reactions from our chemical mechanism should not have a huge impact on the results anyway since the mechanism considers mainly decomposition of alkanes with which $NO_3$ radical reacts to a generally minor extent \citep{Atkinson2008}???.
\item ! the nighttime NO3 radical reaction is of minor significance in term of the overall removal of the alkanes (Atkinson, 1990a)
\item In addition to reacting with organics, $NO_3$ also reacts with $NO_2$, forming dinitrogen pentoxide ($NO_3 + NO_2 \xleftrightarrow{M} N_2O_5$) which in turn is an important nighttime source of nitric acid through its rapid hydrolysis on wet surfaces and aerosol particles ($N_2O_5 + H_2O \xrightarrow{surf., aerosol} 2HNO_3$) \citep{Finlayson-Pitts2000}. Exclusion of $N_2O_5$ from the chemical mechanism assumes no cycling of nitrogen oxides within the troposphere through $NO_3$ and $N_2O_5$, eventually forming $HNO_3$, which are not included into the mechanism anyway.
\item ! In any case, because of this equilibrium, sink for N2O5 such as hydrolysis are, in essence, also sinks for NO3 as well.

@@

\item Another species involved in $NO_x$ tropospheric chemistry is the nitrous acid ($HONO$). The major source of nitrous acid is believed to be heterogeneous 'dark' reactions of $NO_2$, including hydrolysis (reaction with water vapour) of $NO_2$ on aerosol and particulate matter surfaces in nighttime. During the daytime, $HONO$ can be formed by the reaction of $OH$ radical with $NO$ ($OH + NO \xrightarrow{M} HONO$). However, since $HONO$ strongly absorbs UV radiation, it rapidly photolyzes, and under particular conditions (when the nighttime $HONO$ concentration reaches almost 15 ppb \citep{Finlayson-Pitts2000}), photolysis of nighttime-generated $HONO$ can become a major source of $OH$ radicals in the early morning hours ($HONO + hv \rightarrow OH + NO$) \citep{Lammel1996}. Therefore exclusion of $HONO$ from the chemical mechanism potentially leads to underestimation of $OH$ concentrations, but having neither formation nor removal of $HONO$ should balance this effect out???.

@@

\item Inclusion of all chemistry listed above into the model is a subject for future work.

@@

\subsection{Design of experiments}\label{sec:method_expdesign}
To investigate the influence of alkyl nitrate chemistry on (potential changes in) tropospheric $O_3$ production, box model runs were performed at a series of different $NO_x$ and VOCs initial concentrations, so that an $O_3$ isopleth plot could be constructed, similar to those found in \citep{Dodge1977} and \citep{Sillman1999}.   Table \ref{tab:setupO3CO} shows initial concentrations of 'independent'/global parameters and indicates which of them are fixed or allowed to vary during the model run. Table \ref{tab:setupNOx} and Table \ref{tab:setupVOCs} show the initial concentrations of $NO_x$ and VOCs respectively that are used at a series of experiments.

@@

Two experiments have been performed. The aim of the first one is to reveal the total impact of alkyl nitrate chemistry on $O_3$ production, the aim of the second one is to estimate the contribution of individual alkyl nitrates to changes in $O_3$ production (sensitivity!).

For the first experiment the model was run under two configurations, (1) with all alkyl nitrate chemistry switched on, and (2) with all alkyl nitrate chemistry switched off. In every model run VOCs concentrations were (changed according to the data from Table \ref{tab:setupVOCs} and) were fixed throughout the run, while the initial concentrations of $NO_x$ were (changed according to the data from Table \ref{tab:setupNOx} and were) allowed to vary throughout the run. As a result, the first experiment consists of 253 model runs including 11 runs (varying $NO_x$) with inorganic chemistry only.

>>>>

Alkyl nitrates also have a biomass burning source which have been first reported by \cite{Simpson2002}. They found that in the immediate vicinity of savanna fires in Northern Australia the average alkyl nitrate mixing ratios were 47-–122 times higher than local background mixing ratios, exceeding 3 ppbv for methyl nitrate. Methyl nitrate was primarily emitted during the 'flaming' stage, while $C_2$-$C_4$ ethyl--butyl nitrates dominated during the 'smoldering' stage.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\arraystretch}{1}
\begin{table}[h!]
\caption{Mean mixing ratios of the sum of alkyl nitrates ($\Sigma RONO_2$) measured in different environments. The measurement area is indicated in parentheses, type of measurements is in brackets.}\label{tab:ANmean} % mean AN mixing ratios
\begin{tabularx}{\linewidth}{P{3cm}P{2cm}P{5cm}X}
\hline
Environment     & Mean $\Sigma RONO_2$ & $\Sigma RONO_2$ breakdown & Reference \\
\hline
remote marine (North Atlantic) [airborne]   & 10 pptv    & methyl, ethyl, 1-propyl, 2-propyl, 2-butyl, 2-methyl-3-butyl, sum of 2- and 3-pentyl  & \citep{Reeves2007} \\
\hline
rural (Nova Scotia, Canada) [ground]        & 14 pptv    & methyl, ethyl, n-propyl, 2-propyl, 2-butyl  & \citep{Roberts1998} \\
\hline
suburban (Cape Retire, South Africa) [ground] & 17.48 pptv (3.62-122.23) & 2-propyl, 1-propyl, 2-butyl, 1-butyl, 3-pentyl, 2-pentyl, 2-methyl-1-butyl, i-pentyl & \citep{DeKock1994} \\
\hline
suburban (Black Forest, Germany) [ground]   & 119.8 ppt (30-630 ppt) & $C_1-C_8$ & \citep{Flocke1998} \\
\hline
polluted marine (Gulf of Mexico) [airborne] & approx 1 ppb (plume center)& $NO_y-(NO_x+HNO_3+PAN)$ & \citep{Neuman2012} \\
\hline
\end{tabularx}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

A study that enabled us to re-examine the theory of $O_3$ and $RONO_2$ formation processes in a quite unique environment has been conducted by \cite{Neuman2012}. After the Deepwater Horizon (DWH) offshore oil platform accident on 20 April 2010 large amounts of oil from the DWH spill reached the ocean surface. Subsequent oil evaporation caused a release of methane and other light alkanes to the atmosphere forming a $VOC$ mixture atypical for the remote marine environment. At the same the $NO_x$ levels were enhanced in the vicinity of the oil spill due to spill response efforts (ship exhaust, surface oil burning, flaring of recovered gas). Daytime airborne measurements of reactive nitrogen species ($NO_y$ = $NO_x$+$HNO_3$+$PAN$\footnote{Peroxyacetyl nitrate.}+$RONO_2$) and VOCs in the marine boundary layer downwind from the DWH area showed that oxidation of alkanes with between 5 and 10 carbons per molecule ($C_5$–$C_{10}$) dominated ozone formation chemistry in the plume. This circumstance led to a rapid and nearly complete oxidation of $NO_x$ to $RONO_2$ with mostly 6--10 carbons per molecule ($C_6$–$C_{10}$) effectively terminating ozone production. As a result, on 10 June, enhancements ozone levels in the DWH plumes were less than 15 ppbv, even after $NO_x$ was completely oxidized.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h!]
\centering
\includegraphics[width=0.5\linewidth]{{\picspath Neuman2012_Fig.4}.png}
\caption{Time series of reactive nitrogen species measured on 10 June 2010 in the MBL in plumes (a) 10 km and (b) 50 km downwind from DWH. $NO_x$, $NO_y$, and $HNO_3$ were measured once per s and $PAN$ was measured once per 2 s. $C_1$–$C_5$ alkyl nitrates were measured in whole air samples at locations indicated by the circles. The sum of alkyl nitrates (black lines) is inferred from the difference between $NO_y$ and ($NO_x$+$HNO_3$+$PAN$). Source: \citep{Neuman2012}.}\label{fig:Neuman2012}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Estimated $O_3$ production efficiency in the DWH plume is equal to 6$\pm$0.5 ppbv of $O_3$ per ppbv of $NO_x$ oxidized after 1 h of transport. These values are comparable with those measured in U.S. urban areas \citep{Trainer1995,Neuman2009}, but smaller than those measured in the petrochemical plumes with $OH$ loss rates dominated by alkenes (10--18 ppbv $O_3$ per ppbv of $NO_x$) \citep{Ryerson2003}.

Temporal changes in reactive nitrogen partitioning in the DWH plume (Figure \ref{fig:Neuman2012}) nicely illustrate that a quite large concentration of alkyl nitrates in the air can be build up under certain conditions. While in fresh part of the plume (Figure \ref{fig:Neuman2012}a) $NO_x$ accounted for over 80\% of plume $NO_y$, in aged plumes (Figure \ref{fig:Neuman2012}b) $NO_x$ oxidation was nearly complete, but $NO_x$, $PAN$, $HNO_3$, and the sum of eight $C_5$–$C_{10}$ $RONO_2$ accounted for only 22\% of plume $NO_y$. It means that $C_6$–$C_{10}$ $RONO_2$ accounted for 78\% of the aged plume $NO_y$ \citep{Neuman2012}. This result differs dramatically from other studies, where the estimated contribution of $RONO_2$ to $NO_y$ was much lower: approximately 4\% in urban and petrochemical plumes \citep{Ryerson2003}, 10\% in rural areas \citep{Parrish1993}, a few percent in urban air masses \citep{Ryerson2003}, and up to 10–20\% of $NO_y$ in some studies \citep{Day2003}.

Probably the most detailed investigation of the formation processes and photochemical evolution of alkyl nitrates (but not their impact on ozone) has been done by \cite{Sommariva2008}. Using a subset of the Master Chemical Mechanism v3.1 (MCM) they constructed a highly detailed chemical model and run it for 5 days under two scenarios (Table \ref{tab:Sommariva2008init}) to explore the sensitivity of the evolution of the chemical composition of an air mass emitted from an urban environment to the initial conditions. As it seen in Figure \ref{fig:Sommariva2008fig}, the modelling shows that $O_3$ is produced and $NO_x$ is removed from the air mass with the consequent formation of $HNO_3$. In scenario A (Figure \ref{fig:Sommariva2008fig} left) the concentration of $NO_x$ after the first day is about 0.5–-1.0 ppb for $NO_2$ and approximately 100 ppt for $NO$, which is high enough to sustain the production of $O_3$ to up to 135 ppb and the formation of alkyl nitrates during the simulation. In scenario B (Figure \ref{fig:Sommariva2008fig} right) the increase in $O_3$ is slower (note that initial concentration is zero) and the maximum concentration after five days is about 100 ppb. The concentration of $NO_x$ remains relatively high during the second (10–-15 ppb) and third day (3-–5 ppb) and then decreases to $<$200 ppt of $NO$ and $<$1ppb of $NO_2$ on the fourth and fifth days, therefore allowing the formation of alkyl nitrates too. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
\caption{Initial values of some model parameters (in ppb) for two scenarios used in \citep{Sommariva2008}}\label{tab:Sommariva2008init}
\centering
\begin{tabular}{ccc}
\hline
Parameter & Scenario A & Scenario B \\
\hline
$CO$      & 332.5 & 232.5 \\
$NO_x$    & 34.9  & 46.5  \\
$O_3$     & 90    & 0   \\
\hline
\end{tabular}
\end{table}
\begin{figure}
\includegraphics[width=\linewidth]{{D:/UEA/ENV-MB4X_Dissertation/ENV-MB4X_Dissertation_thesis/ENV-MB4X_Dissertation_thesis_latex/pics/Sommariva2008_Fig.1}.png}
\caption{Modelled concentrations of $O_3$, $NO_x$ and $HNO_3$ for scenario A (left) and B (right). Source: \citep{Sommariva2008}.}\label{fig:Sommariva2008fig}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Comparison of modelling results with the values observed inside a urban plume sampled using the NOAA WP-3D aircraft from the New York City area during the New England Air Quality Study (NEAQS) 2004 campaign enabled \cite{Sommariva2008} to conclude that the chemistry of alkyl nitrates is well represented in the MCM model. They found that short-chain $<C_4$ alkyl nitrates can be formed from several precursors, while long-chain ($C_5$) alkyl nitrates are produced for 90\% or more from the oxidation of a single parent alkane. They also found that at lower $O_3$ and $CO$ initial concentrations and higher $NOx$ (scenario A) the importance of the reactions between alkane and $OH$ is typically lower during the first two-three days, and the contributions of peroxides and carboxylic acids become significant later and are on average less important.

The first 3-D chemical transport  model simulations examining the impact of methyl and ethyl nitrates on tropospheric ozone chemistry in remote regions have been conducted by \cite{Neu2008}. In their study, they incorporated methyl and ethyl nitrates in the chemical mechanism of the UCI CTM model \citep{Carver1997}, and by matching their observed atmospheric abundances derived a total oceanic emission flux of these two nitrates (0.35 Tg of N per year). Comparing two simulations with and without spatially uniform emissions of methyl and ethyl nitrates they found that the largest impact of oceanic alkyl nitrates is seen over the equatorial Western Pacific (Figure \ref{fig:Neu2008}). Atmospheric circulation (convergence of the trade winds) in this region isolates the marine boundary layer, resulting in sustained photochemical loss of $O_3$ (Figure \ref{fig:Neu2008M_O3contol}) and at the same time accumulation of alkyl nitrates (Figure \ref{fig:Neu2008M_MeONO2}), which drive up boundary layer $NO_x$ abundances by as much as 250\% (Figure \ref{fig:Neu2008M_NOxdiff}). Such combination of factors leads to increases in $O_3$ of up to 20\% (Figure \ref{fig:Neu2008M_O3diff}), with the largest increases occurring from May to October.

@@

To sum up, alkyl nitrates are produced as a minor branch of the reaction of peroxy radicals with NO \citep{Day2003}. Since these compounds are removed from the troposphere by photolysis and OH oxidation they can provide information on photochemical evolution of an air mass through relationships with their parent hydrocarbons \citep{Bertman1995}. Acting as reservoir species for reactive nitrogen and having relatively long atmospheric lifetimes alkyl nitrates play an important role in long range transport of NOx. Despite their importance and quite large build up under certain conditions alkyl nitrates are usually not included in climate-chemistry models, for example in UKCA (Reeves, personal communication), which potentially could lead to discrepancies between models and observations mentioned before.

\section{Methodology} \label{sec:method}
\subsection{Model development and description}\label{sec:method_develop}
A zero-dimensional (box) model is developed to investigate the impact of alkyl nitrate chemistry on ozone production. Its code is written using a new to the author program \href{http://www.mcpa-software.com/}{FACSIMILE for Windows}. FACSIMILE is a special high-level programming language which is a powerful modelling tool designed to solve differential equations including those resulting from a set of chemical reactions using the Gear’s method.

The model is intended to represent the global troposphere under mean summer daytime meteorology and clear skies. To achieve that a number of assumptions have been made. First, the model assumes a uniform mixing of individual constituents of the troposphere, second, there are no atmospheric transport (no 'winds') and no emissions of species into the box meaning that the time evolution of each species is controlled by the chemical interactions alone and determined by their initial concentrations. The air temperature and relative humidity are assumed to be $25^{\circ}C$ and 70\% respectively and are fixed in all model runs. Other assumptions concerning the chemical mechanism and photolysis rates can be found in the following sections.

@@

\item the formation of alkyl peroxy radicals ($RO_2$) from oxidation of the hydrocarbons;

@@

The full list of reactions incorporated into the code of the model can be found in Appendix \ref{sec:appendix1} and also on this on-line repository: \href{https://github.com/justmeteomary4/ENV-MB4X_model}{\textbf{link}}. It is important to mention that due to purely theoretical purpose of this research and its focus on alkyl nitrate chemistry and ozone some of the reactions that lead to formation of species that serve as a sink for $NO_x$ are not included into the chemical mechanism. However, these reactions play an important role in the chemistry of the troposphere, and below here is a description of their role and anticipated effect of their exclusion.

@@

\item The formation of the nitric acid ($HNO_3$) is a major loss process for $NO_x$ during daytime in the atmosphere. Being a 'sticky' molecule $HNO_3$ readily adsorbs to surfaces, particularly if there is water on the surface, and deposits either in precipitation (wet deposition) or in dry form (dry deposition). Because of the high deposition velocity, dry deposition of $HNO_3$ can be responsible for much of the removal of inorganic nitrogen from the troposphere \citep{Finlayson-Pitts2000}, which means that exclusion of $HNO_3$ chemistry from the chemical mechanism potentially leads to overproduction of ozone in the model, especially at high $NO_2$ concentrations because the competition between $NO_2$ photolysis and its reaction with $OH$ radical ($NO_2 + OH \xrightarrow{M} HNO_3$) is ignored.
\item Since this research is focused on the daytime chemistry some reactions that are believed to be significant in nighttime are not included into the chemical mechanism. Among them are reactions that involve the nitrate radical ($NO_3$) and dinitrogen pentoxide ($N_2O_5$).

@@

\item $NO_3$ radical rapidly photolyzes during the daylight hours, but in nighttime its concentration can increase to measurable levels \citep{Atkinson2008}. This increase in concentration and a lesser availability of the OH radical (which sources are mostly photolytic in nature) make $NO_3$ radical a major contributor to the night chemistry of organics in the troposphere \citep{Finlayson-Pitts2000}. However, exclusion of $NO_3$ radical reactions from our chemical mechanism should not have had a huge impact on the results anyway since the mechanism considers mainly decomposition of alkanes with which $NO_3$ radical reacts to a generally minor extent \citep{Atkinson2008}.
\item In addition to reacting with organics, $NO_3$ also reacts with $NO_2$, forming dinitrogen pentoxide ($NO_3 + NO_2 \xleftrightarrow{M} N_2O_5$) which in turn is an important nighttime source of nitric acid through its rapid hydrolysis on wet surfaces and aerosol particles ($N_2O_5 + H_2O \xrightarrow{surf., aerosol} 2HNO_3$) \citep{Finlayson-Pitts2000}. Exclusion of $N_2O_5$ from the chemical mechanism assumes no cycling of nitrogen oxides within the troposphere through $NO_3$ and $N_2O_5$, eventually forming $HNO_3$, which is not included into the mechanism anyway.

@@

\item Another species involved in $NO_x$ tropospheric chemistry is the nitrous acid ($HONO$). The major source of nitrous acid is believed to be heterogeneous 'dark' reactions of $NO_2$, including hydrolysis (reaction with water vapour) of $NO_2$ on aerosol and particulate matter surfaces in nighttime. During the daytime, $HONO$ can be formed by the reaction of $OH$ radical with $NO$ ($OH + NO \xrightarrow{M} HONO$). However, since $HONO$ strongly absorbs UV radiation, it rapidly photolyzes, and under particular conditions (when the nighttime $HONO$ concentration reaches almost 15 ppb \citep{Finlayson-Pitts2000}), photolysis of nighttime-generated $HONO$ can become a major source of $OH$ radicals in the early morning hours ($HONO + hv \rightarrow OH + NO$) \citep{Lammel1996}. Therefore exclusion of $HONO$ from the chemical mechanism potentially leads to underestimation of $OH$ concentrations, but having neither formation nor removal of $HONO$ should balance this effect out.

@@

@@

\subsection{Design of experiments}\label{sec:method_expdesign}
To investigate the influence of alkyl nitrate chemistry on tropospheric $O_3$ production, box model runs were performed at a series of different $NO_x$ and VOCs initial concentrations, so that an $O_3$ isopleth plot could be constructed, similar to those found in \citep{Dodge1977} and \citep{Sillman1999}.   Table \ref{tab:setupO3CO} shows initial concentrations of main parameters and indicates which of them are fixed or allowed to vary during the model run. Table \ref{tab:setupNOx} and Table \ref{tab:setupVOCs} show the initial concentrations of $NO_x$ and VOCs respectively that are used at a series of experiments.

@@

Two series of experiments have been performed. The aim of the first one is to reveal the total impact of alkyl nitrate chemistry on $O_3$ production, the aim of the second one is to estimate the contribution of individual alkyl nitrates to changes in $O_3$ production.

For the first series of experiments the model was run under two configurations, (1) with all alkyl nitrate chemistry switched on, and (2) with all alkyl nitrate chemistry switched off. In every model run VOCs concentrations were (changed according to the data from Table \ref{tab:setupVOCs}) and were fixed throughout the run, while the initial concentrations of $NO_x$ (Table \ref{tab:setupNOx}) were allowed to vary throughout the run. As a result, the first experiment consists of 253 model runs including 11 runs (with varying $NO_x$) with inorganic chemistry only.

For the second series of experiments the model was run with $NO_x$ initial concentration equal to 10 ppb and VOCs concentrations from case F. In every model run reactions of individual alkyl nitrates or their groups (defined in Table \ref{tab:RHandANs}) were either switched on or switched off, resulting in other 7 additional model runs.
